---
title: "Benchmark"
author: "Zuguang Gu ( guzuguang@suat-sz.edu.cn )"
date: '`r Sys.Date()`'
output: html_vignette
---

```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.width = 6,
    fig.height = 6,
    fig.align = "center"
)
```

## Prepare

The function `unique_strings()` generates a vector of unique strings. It simply uses **digest** package
to generate MD5 hashes of R objects (in the following example, hashed R objects are integers).

```{r}
library(hashtable)
library(microbenchmark)
library(digest)
unique_strings = function(k) {
	sapply(seq_len(k), digest)
}
unique_strings(2)
```

`bm()` is only for simplifying code demo in this document. 

```{r}
bm = function(...) {
	mean(microbenchmark(..., times = 50)$time)
}
```

There are the following packages implementing hash tables:

- **hash**, using environment,
- **hashmapR**, using C++ library `std::unordered_map`,
- **r2r**, using environment.

There are also other indirect hash table implementations:

- `list2env()`, converting a list to an environment.
- **fastmatch**, matching between two character vectors where a hash table is internally
  computated and attached to one character vector.

We compare the performance of creating, querying, deleting and inserting keys 
of **hashtable** and other methods.

## Create hash tables

There are three implementations of hash tables in **hashtable**:

- `hash_table()`, using C++ library `std::unordered_map`,
- `hash_fm_table()`, using **fastmatch**,
- `hash_env_table()`, using environment.

We test hash tables with size from 10K keys to 100K keys.

```{r}
size = seq(10000L, 100000L, by = 10000L)
```

`list2env()` expects the input as a named list. `hashmapR::hashmap()` expects keys and values
both specified as lists. `r2r::hashmap()` creates the hash table in two steps: create an empty
object then fill key-value pairs. Note format preparation is not included in estimating runtime
of benchmarking functions.


```{r, echo = FALSE}
size = c(1000, 5000, size)
# for unknown reasons, the runtime estimates for the first two sizes 
# are not accurate, so we add two sizes to `size` and remove them when making matplot
matplot = function(x, y, ...) {
	graphics::matplot(x[-(1:2)], y[-(1:2), , drop = FALSE], ...)
}
```

```{r}
t1 = t2 = t3 = t4 = t5 = t6 = t7 = numeric(length(size))
for(i in seq_along(size)) {
	keys = unique_strings(size[i])
	values = seq_len(size[i])

	t1[i] = bm(hash_table(keys = keys, values = values))
	t2[i] = bm(hash_fm_table(keys = keys, values = values))
	t3[i] = bm(hash_env_table(keys = keys, values = values))
	
	lt = split(values, keys)
	t4[i] = bm(list2env(lt, hash = TRUE, parent = emptyenv()))
	t5[i] = bm(hash::hash(keys = keys, values = values))

	lt_keys = split(keys, keys)
	lt_values = split(values, keys)
	t6[i] = bm({h = hashmapR::hashmap(); h$set(lt_keys, lt_values, vectorize = TRUE)})
	t7[i] = bm({h = r2r::hashmap(); h[keys] = values})
}
matplot(size, cbind(t1, t2, t3, t4, t5, t6, t7)/1000, type = "o", 
	lty = 1, col = 1:7, pch = 16, cex = 0.5, 
	xlab = "size", ylab = "microseconds", main = "Create hash tables")
legend("topleft", lty = 1, col = 1:7, 
	legend = c("hash_table", "hash_fm_table", "hash_env_table", "list2env", "hash", "hashmapR", "r2r"))
```

Obviously, **hashtable** is the fastest to create hash tables.

## Query

Query a single key. Except **hashmapR**, all hash table objects created by various methods
support `[[` index to get a single value.

```{r}
t1 = t2 = t3 = t4 = t5 = t6 = t7 = numeric(length(size))
for(i in seq_along(size)) {
	keys = unique_strings(size[i])
	values = seq_along(keys)

	h1 = hash_table(keys = keys, values = values)
	h2 = hash_fm_table(keys = keys, values = values)
	h3 = hash_env_table(keys = keys, values = values)
	
	lt = split(values, keys)
	h4 = list2env(lt, hash = TRUE, parent = emptyenv())

	h5 = hash::hash(keys = keys, values = values)
	
	h6 = hashmapR::hashmap()
	h6$set(split(keys, keys), split(values, keys), vectorize = TRUE)

	h7 = r2r::hashmap()
	h7[keys] = values

	t1[i] = bm(h1[[sample(keys, 1)]])
	t2[i] = bm(h2[[sample(keys, 1)]])
	t3[i] = bm(h3[[sample(keys, 1)]])
	t4[i] = bm(h4[[sample(keys, 1)]])
	t5[i] = bm(h5[[sample(keys, 1)]])
	t6[i] = bm(h6$get(sample(keys, 1)))
	t7[i] = bm(h7[[sample(keys, 1)]])
}
matplot(size, cbind(t1, t2, t3, t4, t5, t6, t7)/1000, type = "o", 
	lty = 1, col = 1:7, pch = 16, cex = 0.5, 
	xlab = "size", ylab = "microseconds", main = "Query a single key")
legend("topleft", lty = 1, col = 1:7, 
	legend = c("hash_table", "hash_fm_table", "hash_env_table", "list2env", "hash", "hashmapR", "r2r"))
```

All hash-functions have similar performance. The small difference of runtime
is due to their specific implementations, e.g., **hashtable** implements functions using S4, so 
it needs a little bit more steps on method dispatch.

Query multiple keys. We use `mget()` to get multiple values from an environment. Many
methods support `[` index to get multiple values.

```{r, fig.width = 10, fig.height = 10}
par(mfrow = c(2, 2))
for(p in c(0.01, 0.1, 0.25, 0.5)) {
	t1 = t2 = t3 = t4 = t5 = t6 = t7 = numeric(length(size))
	for(i in seq_along(size)) {
		keys = unique_strings(size[i])
		values = seq_along(keys)

		h1 = hash_table(keys = keys, values = values)
		h2 = hash_fm_table(keys = keys, values = values)
		h3 = hash_env_table(keys = keys, values = values)
		
		lt = split(values, keys)
		h4 = list2env(lt, hash = TRUE, parent = emptyenv())

		h5 = hash::hash(keys = keys, values = values)
		
		h6 = hashmapR::hashmap()
		h6$set(split(keys, keys), split(values, keys), vectorize = TRUE)

		h7 = r2r::hashmap()
		h7[keys] = values

		nk = round(p*size[i])
		t1[i] = bm(h1[sample(keys, nk)])
		t2[i] = bm(h2[sample(keys, nk)])
		t3[i] = bm(h3[sample(keys, nk)])
		t4[i] = bm(mget(sample(keys, nk), h4))
		t5[i] = bm(h5[sample(keys, nk)])
		t6[i] = bm(h6$get(sample(keys, nk)))
		t7[i] = bm(h7[sample(keys, nk)])
	}
	matplot(size, cbind(t1, t2, t3, t4, t5, t6, t7)/1000, type = "o", 
		lty = 1, col = 1:7, pch = 16, cex = 0.5, 
		xlab = "size", ylab = "microseconds", main = paste0("Query multiple keys, ", p*100, "%"))
	legend("topleft", lty = 1, col = 1:7, 
		legend = c("hash_table", "hash_fm_table", "hash_env_table", "list2env", "hash", "hashmapR", "r2r"))
}
```

**r2r** and **hash** performs badly when querying multiple keys. Remove `r2r`,
  `hash()` also remove two environment-functions `list2env()` and
  `hash_env_table()`, then remake the last plot:

```{r}
matplot(size, cbind(t1, t2, t6)/1000, type = "o", 
	lty = 1, col = c(1:2, 6), pch = 16, cex = 0.5, 
	xlab = "size", ylab = "microseconds", main = paste0("Query multiple keys, ", p*100, "%"))
legend("topleft", lty = 1, col = c(1:2, 6), 
	legend = c("hash_table", "hash_fm_table", "hashmapR"))
```

`hash_table()`, `hash_fm_table()` and **hashmapR** only have very small difference.

## Insertion

Except **hashmapR**, all hash table objects created by various methods
support `[[<-` to set a single value. Note `hash_fm_table()` is not incluced
because it does not allow to insert new keys.

Note `unique_strings()` is based on MD5 hashes of integers, `rnorm(1)` generates
numeric values, so `digest(rnorm(1))` generates new keys.

```{r}
t1 = t3 = t4 = t5 = t6 = t7 = numeric(length(size))
for(i in seq_along(size)) {
	keys = unique_strings(size[i])
	values = seq_along(keys)

	h1 = hash_table(keys = keys, values = values)
	h3 = hash_env_table(keys = keys, values = values)
	
	lt = split(values, keys)
	h4 = list2env(lt, hash = TRUE, parent = emptyenv())

	h5 = hash::hash(keys = keys, values = values)
	
	h6 = hashmapR::hashmap()
	h6$set(split(keys, keys), split(values, keys), vectorize = TRUE)

	h7 = r2r::hashmap()
	h7[keys] = values
	
	t1[i] = bm(h1[[digest(rnorm(1))]] <- 1L)
	t3[i] = bm(h3[[digest(rnorm(1))]] <- 1L)
	t4[i] = bm(h4[[digest(rnorm(1))]] <- 1L)
	t5[i] = bm(h5[[digest(rnorm(1))]] <- 1L)
	t6[i] = bm(h6$set(digest(rnorm(1)), 1L))
	t7[i] = bm(h7[[digest(rnorm(1))]] <- 1L)
}
matplot(size, cbind(t1, t3, t4, t5, t6, t7)/1000, type = "o", 
	lty = 1, col = c(1, 3:7), pch = 16, cex = 0.5, 
	xlab = "size", ylab = "microseconds", main = "Insert a single key")
legend("topleft", lty = 1, col = c(1, 3:7), 
	legend = c("hash_table", "hash_env_table", "list2env", "hash", "hashmapR", "r2r"))
```

## Deletion

Difference methods have different interfaces to delete key-value pairs. `hash_fm_table()` is not incluced
because it does not allow to delete keys.

```{r}
t1 = t3 = t4 = t5 = t6 = t7 = numeric(length(size))
for(i in seq_along(size)) {
	keys = unique_strings(size[i])
	values = seq_along(keys)

	h1 = hash_table(keys = keys, values = values)
	h3 = hash_env_table(keys = keys, values = values)
	
	lt = split(values, keys)
	h4 = list2env(lt, hash = TRUE, parent = emptyenv())

	h5 = hash::hash(keys = keys, values = values)
	
	h6 = hashmapR::hashmap()
	h6$set(split(keys, keys), split(values, keys), vectorize = TRUE)

	h7 = r2r::hashmap()
	h7[keys] = values
	
	t1[i] = bm(hash_delete(h1, sample(keys, 1)))
	t3[i] = bm(hash_delete(h3, sample(keys, 1)))
	t4[i] = bm(rm(list = sample(keys, 1), h4))
	t5[i] = bm(hash::delete(sample(keys, 1), h5))
	t6[i] = bm(h6$remove(sample(keys, 1)))
	t7[i] = bm(r2r::delete(h7, sample(keys, 1)))
}
matplot(size, cbind(t1, t3, t4, t5, t6, t7)/1000, type = "o", 
	lty = 1, col = c(1, 3:7), pch = 16, cex = 0.5, 
	xlab = "size", ylab = "microseconds", main = "Delete a single key")
legend("left", lty = 1, col = c(1, 3:7), 
	legend = c("hash_table", "hash_env_table", "list2env", "hash", "hashmapR", "r2r"))
```

`list2env()` performs obviously bad. Note for environment, we did not use `h4[[key]] = NULL` because
it just assign `NULL` to `key` but the key still exists. 

## Compare to match-family functions

People may think named vectors (including lists) behave like hashes.

```{r}
vec = c("foo" = 1, "bar" = 2)
vec
vec[["foo"]]

lt = list("foo" = 1, "bar" = "text")
lt$foo
```

Actually they are not. Names are just normal character vectors, but attached to the object as an attributes.

```{r}
attributes(vec)
attributes(lt)
```

So using names as indices basically does matching in the "names" attributes.

There are following match-family functions:

- `pmatch()`, used in `names()`, `$`, `[[`, `[`,
- `match()`, used in `%in%`, `intersect()`, `unique()`, `duplicated()`,
- **fastmatch**, a faster version of `match()`, synonyms functions are `fmatch()` and `%fin%`,
- `data.table::chmatch()`, also a faster version of `match()`, synonyms functions are `chmatch()` and `%chin%`.

`pmatch()` means partial matching. It matches elements if substrings can be uniquely mapped.

`match()`, **fastmatch** and `chmatch()` do the complete matching. They have the same interface：

```
function(x, table, ...)
```

where `x` is the query vector, i.e. the keys, and `table` is the data vector, i.e. all keys.
First, all of `match()`, **fastmatch** and `chmatch()` internally generate hash tables for fast matching.
The difference is `match()` and `chmatch()` generate hash tables in every function call, which means, 
even `table` is the same, multiple calls to `match()` or `chmatch()`, the internal hash table needs to 
be rebuilt, while **fastmatch** directly attaches the hash table to `table` as an internal attribute,
so in afterward repeated calls of `fastmatch::fmatch()`, the hash table can be directly reused.


```{r}
library(fastmatch)
library(data.table)
t1 = t2 = t3 = t4 = t5 = t6 = t7 = t8 = numeric(length(size))
for(i in seq_along(size)) {
	keys = unique_strings(size[i])
	values = seq_along(keys)

	x = structure(values, names = keys)

	h1 = hash_table(keys = keys, values = values)
	h2 = hash_fm_table(keys = keys, values = values)
	h3 = hash_env_table(keys = keys, values = values)

	t1[i] = bm(h1[[sample(keys, 1)]])
	t2[i] = bm(h2[[sample(keys, 1)]])

	h6 = hashmapR::hashmap()
    h6$set(split(keys, keys), split(values, keys), vectorize = TRUE)
    t6[i] = bm(h6$get(sample(keys, 1)))

	t4[i] = bm(x[[sample(keys, 1)]])
	t5[i] = bm(x[[match(sample(keys, 1), keys)]])
	t7[i] = bm(x[[fmatch(sample(keys, 1), keys)]])
	t8[i] = bm(x[[chmatch(sample(keys, 1), keys)]])
}
matplot(size, cbind(t1, t2, t3, t6, t4, t5, t7, t8)/1000, type = "o", 
	lty = 1, col = c(1:3, 6, 4, 5, 7, 8), pch = 16, cex = 0.5, 
	xlab = "size", ylab = "microseconds", main = "Query a single key, match-family")
legend("topleft", lty = 1, col = c(1:3, 6, 4, 5, 7, 8), 
	legend = c("hash_table", "hash_fm_table", "hash_env_table", "hashmapR", "pmatch/names", "match", "fastmatch", "chmatch"))
```

We can see hash-family functions have time complexity of $O(1)$ to the hash
table size once the hash table is already created.
**fastmatch** is also fast because the hash table only needs to be calculated once.
`pmatch()`, `match()` and `chmatch()` have linear complexity to the size of "the data vector".

Next query multiple keys with match-family functions. 

```{r, fig.width = 10, fig.height = 10}
par(mfrow = c(2, 2))
for(p in c(0.01, 0.1, 0.25, 0.5)) {
	t1 = t2 = t3 = t4 = t5 = t6 = t7 = numeric(length(size))
	for(i in seq_along(size)) {
		keys = unique_strings(size[i])
		values = seq_along(keys)

		x = structure(values, names = keys)

		h1 = hash_table(keys = keys, values = values)
		h2 = hash_fm_table(keys = keys, values = values)
		h3 = hash_env_table(keys = keys, values = values)

		t1[i] = bm(h1[sample(keys, round(p*size[i]))])
		t2[i] = bm(h2[sample(keys, round(p*size[i]))])
		t3[i] = bm(h3[sample(keys, round(p*size[i]))])

		h6 = hashmapR::hashmap()
	    h6$set(split(keys, keys), split(values, keys), vectorize = TRUE)
	    t6[i] = bm(h6$get(sample(keys, round(p*size[i]))))
			
		t4[i] = bm(x[sample(keys, round(p*size[i]))])
		t5[i] = bm(x[match(sample(keys, round(p*size[i])), keys)])
		t7[i] = bm(x[fmatch(sample(keys, round(p*size[i])), keys)])
		t8[i] = bm(x[chmatch(sample(keys, round(p*size[i])), keys)])
	}
	matplot(size, cbind(t1, t2, t3, t6, t4, t5, t7, t8)/1000, type = "o", 
		lty = 1, col = c(1:3, 6, 4, 5, 7, 8), pch = 16, cex = 0.5, 
		xlab = "size", ylab = "microseconds", main = paste0("Query multiple keys, match-family, ", p*100, "%"))
	legend("topleft", lty = 1, col = c(1:3, 6, 4, 5, 7, 8), 
		legend = c("hash_table", "hash_fm_table", "hash_env_table", "hashmapR", "pmatch/names", "match", "fastmatch", "chmatch"))
}
```

Now the match-family functions have good performance when querying multiple keys. As has been explained,
for multiple calls,  match-family functions performs bad because the internal hash tables have to be rebuilt,
but for multiple key queries in a single call, keys can be directly queried from the internal hash table.

Remove `hash_env_table()` and remake the last plot.

```{r}
matplot(size, cbind(t1, t2, t6, t4, t5, t7, t8)/1000, type = "o", 
	lty = 1, col = c(1:2, 6, 4, 5, 7, 8), pch = 16, cex = 0.5, 
	xlab = "size", ylab = "microseconds", main = paste0("Query multiple keys, match-family, ", p*100, "%"))
legend("topleft", lty = 1, col = c(1:2, 6, 4, 5, 7, 8), 
	legend = c("hash_table", "hash_fm_table", "hashmapR", "pmatch/names", "match", "fastmatch", "chmatch"))
```


## Hash set

Hash set is basically very similar as hash table. The only difference is for
hash table, there are values associated with keys, while for hash set, there
are only keys.

We first test querying a single key, which is to test whether the key exists in the hash set.

```{r}
t1 = t2 = t3 = t4 = t5 = t6 = t7 = numeric(length(size))
for(i in seq_along(size)) {
	keys = unique_strings(size[i])

	h1 = hash_set(keys)
	h2 = hash_fm_set(keys)
	h3 = hash_env_set(keys)
	h7 = r2r::hashset(keys)

	t1[i] = bm(h1[[sample(keys, 1)]])
	t2[i] = bm(h2[[sample(keys, 1)]])
	t3[i] = bm(h3[[sample(keys, 1)]])
	t7[i] = bm(h7[[sample(keys, 1)]])

	t4[i] = bm(match(sample(keys, 1), keys))
	t5[i] = bm(fmatch(sample(keys, 1), keys))
	t6[i] = bm(chmatch(sample(keys, 1), keys))
}
matplot(size, cbind(t1, t2, t3, t7, t4, t5, t6)/1000, type = "o", 
	lty = 1, col = c(1:3, 7, 4:6), pch = 16, cex = 0.5, 
	xlab = "size", ylab = "microseconds", main = "Query a single key")
legend("topleft", lty = 1, col = c(1:3, 7, 4:6), 
	legend = c("hash_set", "hash_fm_set", "hash_env_set", "r2r", "match", "fastmatch", "chmatch"))
```

We remove `match()` and `chmatch()` and remake the last plot.

```{r}
matplot(size, cbind(t1, t2, t3, t7, t5)/1000, type = "o", 
	lty = 1, col = c(1:3, 7, 5), pch = 16, cex = 0.5, 
	xlab = "size", ylab = "microseconds", main = "Query a single key")
legend("topleft", lty = 1, col = c(1:3, 7, 5), 
	legend = c("hash_set", "hash_fm_set", "hash_env_set", "r2r", "fastmatch"))
```

Next query multiple keys.

```{r, fig.width = 10, fig.height = 10}
par(mfrow = c(2, 2))
for(p in c(0.01, 0.1, 0.25, 0.5)) {
	t1 = t2 = t3 = t4 = t5 = t6 = t7 = numeric(length(size))
	for(i in seq_along(size)) {
		keys = unique_strings(size[i])

		h1 = hash_set(keys)
		h2 = hash_fm_set(keys)
		h3 = hash_env_set(keys)
		h7 = r2r::hashset(keys)

		t1[i] = bm(h1[sample(keys, round(p*size[i]))])
		t2[i] = bm(h2[sample(keys, round(p*size[i]))])
		t3[i] = bm(h3[sample(keys, round(p*size[i]))])
		t7[i] = bm(h7[sample(keys, round(p*size[i]))])

		t4[i] = bm(match(sample(keys, round(p*size[i])), keys))
		t5[i] = bm(fmatch(sample(keys, round(p*size[i])), keys))
		t6[i] = bm(chmatch(sample(keys, round(p*size[i])), keys))
	}
	matplot(size, cbind(t1, t2, t3, t7, t4, t5, t6)/1000, type = "o", 
		lty = 1, col = c(1:3, 7, 4:5), pch = 16, cex = 0.5, 
		xlab = "size", ylab = "microseconds", main = paste0("Query multiple keys, ", p*100, "%"))
	legend("topleft", lty = 1, col = c(1:3, 7, 4:6), 
		legend = c("hash_set", "hash_fm_set", "hash_env_set", "r2r", "match", "fastmatch", "chmatch"))
}
```

Obviously `r2r` has very bad performance when querying multiple keys. 
Remove `r2r` also `hash_env_set()` and remake the last plot.

```{r}
matplot(size, cbind(t1, t2, t4, t5, t6)/1000, type = "o", 
	lty = 1, col = c(1:2, 4:6), pch = 16, cex = 0.5, 
	xlab = "size", ylab = "microseconds", main = paste0("Query multiple keys, ", p*100, "%"))
legend("topleft", lty = 1, col = c(1:2, 4:6), 
	legend = c("hash_set", "hash_fm_set", "match", "fastmatch", "chmatch"))
```

## Compare hash tables and hash sets

In **hashtable**, hash tables and hash sets are implements similarly in the three implementations.

```{r}
t1 = t2 = t3 = t4 = t5 = t6 = numeric(length(size))
for(i in seq_along(size)) {
    keys = unique_strings(size[i])
    values = seq_len(size[i])

    h1 = hash_table(keys = keys, values = values)
    h2 = hash_fm_table(keys = keys, values = values)
    h3 = hash_env_table(keys = keys, values = values)

    h4 = hash_set(keys = keys)
    h5 = hash_fm_set(keys = keys)
    h6 = hash_env_set(keys = keys)
    
    p = 0.1
    t1[i] = bm(h1[sample(keys, round(p*size[i]))])
    t2[i] = bm(h2[sample(keys, round(p*size[i]))])
    t3[i] = bm(h3[sample(keys, round(p*size[i]))])
    t4[i] = bm(h4[sample(keys, round(p*size[i]))])
    t5[i] = bm(h5[sample(keys, round(p*size[i]))])
    t6[i] = bm(h6[sample(keys, round(p*size[i]))])
}
matplot(size, cbind(t1, t2, t3, t4, t5, t6)/1000, type = "o", 
    lty = c(1, 1, 1, 2, 2, 2), col = c(1, 2, 3, 1, 2, 3), pch = 16, cex = 0.5, 
    xlab = "size", ylab = "microseconds", main = "Query multiple keys")
legend("topleft", lty = c(1, 1, 1, 2, 2, 2), col = c(1, 2, 3, 1, 2, 3), 
    legend = c("hash_table", "hash_fm_table", "hash_env_table", "hash_set", "hash_fm_set", "hash_env_set"))
```


## Summarize

For creating hash tables, **hashtable** is fast.

For querying single keys, all hash-functions are fast, but the runtime 
of `pmatch()`, `match()` and `chmatch()` increases linearly to the size of the data.

For querying multiple keys, `hash_fm_set()` and `fmatch()` is the fastest. If hash
tables are changable, then `hash_table()` and **hashmapR** can be considered as replacement.

For inserting new keys, all hash-functions are similar.

For deleting keys, `list2env()` (which uses `rm()`) is relatively slow. Other functions are similar.


```{r}
sessionInfo()
```
