---
title: "Benchmark on Hash Tables and Hash Sets"
author: "Zuguang Gu ( z.gu@dkfz.de )"
date: '`r Sys.Date()`'
output: html_vignette
---

```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.width = 6,
    fig.height = 6,
    fig.align = "center"
)
```


The function `random_character_vector()` generates random character vectors.

```{r}
library(hashtable)
library(hash)
library(fastmatch)
library(microbenchmark)
random_character_vector = function(k, nc = 8) {
	unique(do.call(paste0, lapply(1:nc, function(i) {
		sample(c(letters, LETTERS), k, replace = TRUE)
	})))
}
```

## Hash tables

We first benchmark various methods for hash tables, i.e. giving keys returns corresponding values. 
The following methods will be tested.

1. If the input variable is a named vector, use names as indicies. It internally applies `pmatch()`.
2. The same scenario as above, but we use `match()` which internally uses a hash table.
3. The same scenario as above, but we use the **fastmatch** to replace `match()`.
4. The same data is converted to a list of hashed environment.
5. The data is converted to a `hash_table` object with the **hashtable** package.
6. The data is converted to a `hash` object with the **hash** package, which internally implements hash with environment.


First, we benchmark the performance of getting a single element. The total set size changes from 5,000 to 100,000.

```{r}
k = seq(5000, 100000, by = 5000)
t1 = t2 = t3 = t4 = t5 = t6 = numeric(length(k))
for(i in seq_along(k)) {
	keys = random_character_vector(k[i])

	# a named vector
	values = seq_along(keys)
	x1 = structure(values, names = keys)

	lt = split(values, keys)
	x3 = list2env(lt, hash = TRUE)

	x4 = hash_table(keys = keys, values = values)

	x5 = hash(keys = keys, values = values)

	# characters as indicies internally calls pmatch()
	t1[i] = mean(microbenchmark(foo = x1[[sample(keys, 1)]], times = 10)$time)
	# match() returns integer indicies
	t2[i] = mean(microbenchmark(foo = x1[[match(sample(keys, 1), keys)]], times = 10)$time)
	# fmatch() also returns integer indicies
	t3[i] = mean(microbenchmark(foo = x1[[fmatch(sample(keys, 1), keys)]], times = 10)$time)
	# `[[` method for the environment object
	t4[i] = mean(microbenchmark(foo = x3[[sample(keys, 1)]], times = 10)$time)
	# `[[` method for the hash_table object
	t5[i] = mean(microbenchmark(foo = x4[[sample(keys, 1)]], times = 10)$time)
	# `[[` method for the hash object
	t6[i] = mean(microbenchmark(foo = x5[[sample(keys, 1)]], times = 10)$time)
}
matplot(k, cbind(t1, t2, t3, t4, t5, t6)/1000, type = "l", lty = 1, col = 1:6, pch = 16, cex = 0.5, ylab = "microseconds")
legend("topleft", lty = 1, col = 1:6, legend = c("names/pmatch", "match", "fastmatch", "list2env", "hash_table", "hash"))
```

Next, we benchmark the performance of getting multiple elements.
In this scenario, the "list2env" method is excluded since it only allows to get one single element at a time.

For each set size `k`, we pick a fraction of `p` elements.

```{r, fig.width = 10, fig.height = 10}
k = seq(1000, 20000, by = 1000)
par(mfrow = c(2, 2))
for(p in c(0.05, 0.1, 0.2, 0.5)) {
	t1 = t2 = t4 = t5 = numeric(length(k))
	for(i in seq_along(k)) {
		keys = random_character_vector(k[i])

		values = seq_along(keys)
		x1 = structure(values, names = keys)

		x4 = hash_table(keys = keys, values = values)

		x5 = hash(keys = keys, values = values)

		k2 = round(p*k[i])
		t1[i] = mean(microbenchmark(foo = x1[sample(keys, k2)], times = 10)$time)
		t2[i] = mean(microbenchmark(foo = x1[match(sample(keys, k2), keys)], times = 10)$time)
		t3[i] = mean(microbenchmark(foo = x1[fmatch(sample(keys, k2), keys)], times = 10)$time)
		t5[i] = mean(microbenchmark(foo = x4[sample(keys, k2)], times = 10)$time)
		t6[i] = mean(microbenchmark(foo = x5[sample(keys, k2)], times = 10)$time)
	}
	matplot(k, cbind(t1, t2, t4, t5, t6)/1000, type = "o", lty = 1, col = c(1, 2, 4, 5, 6), pch = 16, cex = 0.5, main = paste0("p = ", p), ylab = "microseconds", xlab = "size of the set")
	legend("topleft", lty = 1, col = c(1, 2, 4, 5, 6), legend = c("names/pmatch", "match", "fastmatch", "hash_table", "hash"))
}
```

In general, when only getting one single elements, `list2env()`, **hash** and **hashtable** performs better while `pmatch()` and `match()` have
very bad performance. And when getting multiple elements, **hash** has a significantly bad performance, while `fastmatch()` has the best performance.
The performance of `pmatch()` and `match()` is almost identical and it gets better when the number of elements to retrieve increases.


## Hash set

Hash set is basically very similar as hash table. The only difference is for hash table, there are values associated with keys, while
for hash set, there are only keys.

We use a real world example with 20K genes and a list of 15K gene sets, as overlapping gene sets to genes is a common
task in Bioinformatics.

```{r}
load("test.Rdata") # gs and pc_genes
length(pc_genes)
length(gs)
```

First, we test the performance of getting a single element in the global gene list. The size of the global gene list changes from 1,000 to 20,000.
We test `match()`, `fmatch()`, hash table and the hash set. Note `match()` is used for other match functions such as `intersect()` and `%in%` in R.

```{r}
k = seq(1000, 20000, by = 1000)
for(i in seq_along(k)) {
	keys = sample(pc_genes, k[i])

	h1 = hash_set(keys)
	h2 = hash_table(keys, seq_along(keys))

	t1[i] = mean(microbenchmark(foo = match(sample(keys, 1), keys), times = 10)$time)
	t2[i] = mean(microbenchmark(foo = fmatch(sample(keys, 1), keys), times = 10)$time)
	t3[i] = mean(microbenchmark(foo = hash_set_exists(h1, sample(keys, 1)), times = 10)$time)
	t4[i] = mean(microbenchmark(foo = hash_table_exists(h2, sample(keys, 1)), times = 10)$time)
}
matplot(k, cbind(t1, t2, t3, t4)/1000, type = "o", lty = 1, col = 1:4, pch = 16, cex = 0.5, ylab = "microseconds", xlab = "size of the set")
legend("topleft", lty = 1, col = 1:4, legend = c("match", "fastmatch", "hash_set", "hash_table"))
```

Next, we benchmark the performance of overlapping gene sets to the global gene list.

```{r, fig.width = 10, fig.height = 5}
t1 = t2 = t3 = t4 = t5 = numeric(length(k))
for(i in seq_along(k)) {
	keys = sample(pc_genes, k[i])

	h1 = hash_set(keys)
	h2 = hash_table(keys, seq_along(keys))

	# %in% is basically the same as match()
	t1[i] = microbenchmark(foo = lapply(gs, function(x) x[x %in% keys]), times = 1)$time
	# %fin$ is basically the same as fmatch()
	t2[i] = microbenchmark(foo = lapply(gs, function(x) x[x %fin% keys]), times = 1)$time
	t3[i] = microbenchmark(foo = lapply(gs, function(x) x[hash_set_exists(h1, x)]), times = 1)$time
	t4[i] = microbenchmark(foo = lapply(gs, function(x) x[hash_table_exists(h2, x)]), times = 1)$time
}
par(mfrow = c(1, 2))
matplot(k, cbind(t1, t2, t3, t4)/1e6, type = "o", lty = 1, col = 1:5, pch = 16, cex = 0.5, ylab = "miliseconds", xlab = "size of the set")
legend("topleft", lty = 1, col = 1:4, legend = c("match", "fastmatch", "hash_set", "hash_table"))

# we remove the line for `match()`
matplot(k, cbind(t2, t3, t4)/1e6, type = "o", lty = 1, col = 2:4, pch = 16, cex = 0.5, ylab = "miliseconds", xlab = "size of the set", ylim = c(0, max(t4)/1e6*1.05))
legend("topleft", lty = 1, col = 2:4, legend = c("fastmatch", "hash_set", "hash_table"))
```


We can see for matching single element from the set, `hash_set()` and `hash_table()` have the best performance. When matching
to multiple elements from the set, `match()` has the worst performance compared to others. `fmatch()`/`%fin%` is the most efficient.